---
title: "Getting Started with mxmmod"
author: "Kyle Husmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mxmmod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
eval_semplots <- F
```

## Overview

This tutorial introduces the `mxmmod` package, for building Measurement Model of Derivatives (MMOD; Estabrook, 2015) with OpenMx.

## Outline

A. Introduction to the Measurement Model of Derivatives (MMOD)
B. Data set
C. Example 1: One factor model
D. Example 2: Two factor model
E. Discussion

### Prelim: Prepare environment

Before we begin, install the `mxmmod` package if you have not already:

```{r, eval=F}
devtools::install_github('khusmann/mxmmod')
```

Now let's load the required libraries for this tutorial:

```{r, message=F, warning=F}
library(tidyverse)
library(OpenMx)
library(mxmmod)
```

## A. Introduction to the Measurement Model of Derivatives (MMOD)

Describe mmod here, add formulas etc.

## B. Data set

This tutorial will make use of data from the National Longitudinal Survey of Youth. The NLSY97  sample has a 5-item depression scale administered at 

```{r}
data("nlsy97depression")
summary(nlsy97depression)
```

Plot a few example trajectories:

```{r, fig.width=6, fig.height=6, fig.align='center'}
set.seed(1000)
subset <- sample(unique(nlsy97depression$pid), 9)

nlsy97depression %>%
  filter(pid %in% subset) %>%
  gather(measure, val, -pid, -occasion) %>%
  ggplot(aes(x=occasion, group=measure, color=measure, y=val)) +
  geom_line(position=position_jitter(w=0.1, h=0.1)) +
  facet_wrap(~pid)
```

Plot mean trajectories:

```{r}
nlsy97depression %>%
  select(-pid) %>%
  gather(measure, val, -occasion) %>%
  group_by(occasion, measure) %>%
  na.omit() %>%
  ggplot(aes(x=occasion, color=measure, y=val)) +
  stat_summary(fun.y = mean, geom='line') +
  stat_summary(fun.y = mean, geom='point') +
  stat_summary(fun.data = mean_se, geom='errorbar', width=0.2)
```

## C. Example 1: One factor model

```{r}
structure <- list(
  F1 = c('nervous', 'down', 'depressed', 'calm', 'happy')
)
mmod_model <- mxMmodModel(data=nlsy97depression,
                          modelName='1 Factor MMOD',
                          idvar='pid', timevar='occasion', structure=structure, fiml=F)
mmod_fit <- mxRun(mmod_model)
(mmod_summary <- summary(mmod_fit))
```

```{r, eval=eval_semplots, fig.width=6, fig.height=4, fig.align='center'}
# Note: This can take a while to draw...
semPlot::semPaths(mmod_fit, 'est')
```


## D. Example 2: Two factor model

```{r}
structure2 <- list(
  F1 = c('nervous', 'down', 'depressed'),
  F2 = c('happy', 'calm')
)
mmod_model2 <- mxMmodModel(data=nlsy97depression,
                          modelName='2 Factor MMOD',
                          idvar='pid', timevar='occasion', structure=structure2)
mmod_fit2 <- mxRun(mmod_model2)
(mmod_summary2 <- summary(mmod_fit2))
```

```{r, eval=eval_semplots, fig.width=6, fig.height=4, fig.align='center'}
# Note: This can take a while to draw...
semPlot::semPaths(mmod_fit2, 'est')
```

## E. Discussion

```{r}
fits <- list(mmod_summary, mmod_summary2)

(compare_models <- tibble(
    name=map_chr(fits, 'modelName'),
    chisq=map_dbl(fits, 'Chi'),
    dof=map_dbl(fits, 'ChiDoF'),
    `-2ll`=map_dbl(fits, 'Minus2LogLikelihood'),
    aic=map_dbl(fits, 'AIC.Mx'),
    bic=map_dbl(fits, 'BIC.Mx'),
    rmsea=map_dbl(fits, 'RMSEA'),
    cfi=map_dbl(fits, 'CFI'),
    tli=map_dbl(fits, 'TLI')  
))
```
